final-project

### 2024-11-13 final-project.Rmd
make categorical vars

# there are several variables that need to be categorized for analysis and/filtered
seer_cat <- seer_clean %>%
   filter(
    !is.na(subtype),
    #!is.na(grade),
    #n_stage != "NX"
  ) %>%
  mutate(agecat = case_when(
    age %in% c("20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years") ~ 0,
    age %in% c("50-54 years", "55-59 years", "60-64 years", "65-69 years") ~ 1,
    age %in% c("70-74 years", "75-79 years", "80-84 years", "85+ years") ~ 2,
    TRUE ~ NA_real_  # Keeps other values as NA if they don’t match any of the above groups
  )) %>%
  mutate(gradecat = case_when(
    grade %in% c(1, 2) ~ "I/II",
    grade %in% c(3, 4) ~ "III/IV",
    TRUE ~ NA_character_
    )) %>%
  # the survival_status column is what we use for event/censoring
  # recode Dead as 1 and Alive (censored) as 0
  mutate(status_cat = case_when(
    survival_status == "Dead" ~ 1,
    survival_status == "Alive" ~ 0
  ))

# further filter and select only variables used in analysis
seer <- seer_cat %>%
  select(
    age = agecat,
    sex,
    year_diagnosis,
    race,
    marriage_status,
    subtype,
    n_stage,
    m_stage,
    grade = gradecat,
    radiation,
    chemotherapy,
    surgery,
    survival_months,
    survival_status = status_cat,

@@

## age
CrossTable(age, subtype, prop.r = FALSE, prop.c = TRUE, prop.t = FALSE, prop.chisq = FALSE, chisq = TRUE)

@@

## age
summary(survfit(Surv(time = survival_months, event = survival_status) ~ age), times = c(12, 60, 120))
survdiff(Surv(time = survival_months, event = survival_status) ~ age)

@@

```{r}
# check the proportional hazards assumption for each covariate

## age
ggsurvplot(survfit(Surv(time = survival_months, event = survival_status) ~ age), data = seer, fun="cloglog")

@@

>>>>

# there are several variables that need to be categorized for analysis and/filtered
seer_cat <- seer_clean %>%
   filter(
    !is.na(subtype),
    #!is.na(grade),
    #n_stage != "NX"
  ) %>%
  mutate(agecat = case_when(
    age %in% c("20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years") ~ 0,
    age %in% c("50-54 years", "55-59 years", "60-64 years", "65-69 years") ~ 1,
    age %in% c("70-74 years", "75-79 years", "80-84 years", "85+ years") ~ 2,
    TRUE ~ NA_real_  # Keeps other values as NA if they don’t match any of the above groups
  )) %>%
  mutate(agecat0 = case_when(
    agecat == 0 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(agecat1 = case_when(
    agecat == 1 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(agecat2 = case_when(
    agecat == 2 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat0 = case_when(
    race == "White" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat1 = case_when(
    race == "Black" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(racecat2 = case_when(
    race == "Other" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(marriage_status = case_when(
    marriage_status == "Yes" ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtypecat = case_when(
    subtype == "Luminal A" ~ 0,
    subtype == "Luminal B" ~ 1,
    subtype == "Triple Negative" ~ 2,
    subtype == "HER2 Positive" ~ 3,
    TRUE ~ NA_integer_
  )) %>%
  mutate(subtype0 = case_when(
    subtypecat == 0 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype1 = case_when(
    subtypecat == 1 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype2 = case_when(
    subtypecat == 2 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(subtype3 = case_when(
    subtypecat == 3 ~ 1,
    TRUE ~ 0
  )) %>%
  mutate(gradecat = case_when(
    grade %in% c(1, 2) ~ 0,
    grade %in% c(3, 4) ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(n_stage = case_when(
    n_stage %in% c("N0", "N1") ~ 0,
    n_stage %in% c("N2", "N3") ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(m_stage = case_when(
    m_stage == "M0" ~ 0,
    m_stage == "M1" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(radiation = case_when(
    radiation == "No/Unknown" ~ 0,
    radiation == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(chemotherapy = case_when(
    chemotherapy == "No/Unknown" ~ 0,
    chemotherapy == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  mutate(surgery = case_when(
    surgery == "No/Unknown" ~ 0,
    surgery == "Yes" ~ 1,
    TRUE ~ NA_integer_
    )) %>%
  # the survival_status column is what we use for event/censoring
  # recode Dead as 1 and Alive (censored) as 0
  mutate(status_cat = case_when(
    survival_status == "Dead" ~ 1,
    survival_status == "Alive" ~ 0
  ))

# further filter and select only variables used in analysis
seer <- seer_cat %>%
  select(
    agecat,
    agecat0,
    agecat1,
    agecat2,
    sex,
    year_diagnosis,
    race,
    racecat0,
    racecat1,
    racecat2,
    marriage_status,
    subtype = subtypecat,
    subtype0,
    subtype1,
    subtype2,
    subtype3,
    n_stage,
    m_stage,
    grade = gradecat,
    radiation,
    chemotherapy,
    surgery,
    survival_months,
    survival_status = status_cat,

@@

## agecat
CrossTable(agecat, subtype, prop.r = FALSE, prop.c = TRUE, prop.t = FALSE, prop.chisq = FALSE, chisq = TRUE)

@@

## agecat
summary(survfit(Surv(time = survival_months, event = survival_status) ~ agecat), times = c(12, 60, 120))
survdiff(Surv(time = survival_months, event = survival_status) ~ agecat)

@@

```{r}
# check the proportional hazards assumption for each covariate with (1) a log log plot

## agecat
ggsurvplot(survfit(Surv(time = survival_months, event = survival_status) ~ agecat), data = seer, fun="cloglog")

@@

```{r}
# create a dataset in counting process format
seer.cp <- survSplit(seer, cut=seer$survival_months[seer$survival_status==1],
                     end="survival_months",event="survival_status", start="start")

```

```{r}
# check PH assumption for race, n_stage, chemotherapy, surgery with statistical approach

# create time-varying covariates for race by adding adding a product term with the ln of time
seer.cp$logtrace1=seer.cp$race[seer.cp$race == "White"]*log(seer.cp$survival_months)

seer.cp$logtrace <- ifelse(seer.cp$race == "White", log(seer.cp$survival_months), NA)

```

```{r}
model <- coxph(
  Surv(start, survival_months, survival_status) ~ subtype0 + subtype1 + subtype2 + subtype3 +  
    agecat0 + agecat1 + agecat2 + 
    racecat0 + racecat1 + racecat2 +
    marriage_status +

               factor(SEX) + factor(cholcat) + CURSMOKE + bmicat1 + bmicat2, data = fham.cp3)
```
